Theories of deliberated choice

### 2023-11-10 simple.tex
Rm outdated cmted

Given a pair $(I, \allargs)$, an observation $θ \in \Theta$ is a right unique finite subset of $(I × \allhists) × P$ such that $\forall i \in I, (i, \hist, p), (i, \histp, q) \in θ: \card{\hist} ≤ \card{\hist'} \implies \hist \subseteq \hist'$.
The set $\Theta$ contains the a priori possible observations, that is, those that can be considered possible when nothing is known about $\allleadsto$.
% Let $S_θ = \set{{\allleadstog} \in \allleadstos \suchthat θ \subseteq {\allleadstog}}$ denote the possibilities for $\allleadsto$ compatible with the observations.
We say that an observation $\theta$ is compatible with a claim $C$ iff
% $S_θ \cap C ≠ \emptyset$, equivalently, iff 
$\theta \in C^\subseteq$, equivalently, iff
$\theta \in \cup_{{\allleadstog} \in C} \powerset{\allleadstog}$.
Thus, given ${\allleadstog} \in \allleadstos$, the set $\Theta \cap \powerset{\allleadstog}$ is the set of observations compatible with ${\allleadstog}$.

@@

Let the \emph{simple a priori observations} be defined as $\cup_{J \subseteq_\finite I} (\allhists)^J$, equivalently, as the right-unique finite relations over $I × \allhists$.
Given a simple a priori observation $O: J \mapsto \allhists$, we can define the a priori observation $O_O = \set{(i, \hist) \in J × \allhists \suchthat \forall i \in J: \hist \subseteq O(i)}$ as the completion of $O$ to all sub-paths in $O$.
Given $O$ and ${\allleadstog}$, it follows that $S_{\restr{\allleadstog}{O_O}}$ denotes the “leads to” relations that complete $O_O$ using $\allleadstog$.

@@

At our disposal is the procedure of empirical rejection, that consists in directly invalidating a claim by observing something that contradicts it.
\begin{definition}
  Given an exposure protocol $\allleadsto$, we say that $C$ is \emph{rejectable} iff
  % $\exists θ \in \Theta \cap \powerset{\allleadsto} \suchthat θ \notin C^\wsub$.
  $\Theta \cap \powerset{\allleadsto} \nsubseteq C^\wsub$.
\end{definition}

The following theorem verifies that when $C$ holds, it is not rejectable.
\begin{theorem}
  \label{th:ex-post}
  $\forall C \ni {\allleadsto}:
    % \nexists θ \in \Theta \suchthat \theta \subseteq {\allleadsto} \land θ \notin C^\wsub$.
    \Theta \cap \powerset{\allleadsto} \subseteq C^\wsub$.
\end{theorem}
\begin{proof}
  This holds by definition of $C^\wsub = \cup_{{\allleadstog} \in C} \powerset{\allleadstog}$.
  % Consider any claim $C$ with ${\allleadsto} \in C$.
  % Consider any $\theta \in \Theta$ with $\theta \wsub {\allleadsto}$.
  % We have $θ \subseteq {\allleadsto} \in C$, thus $θ \in C^\wsub$.
\end{proof}

The converse, however, does not hold.
\begin{example}[Disjunct with incompatible paths, not rejectable]
  \label{ex:disjR}
  Given $\ar \neq \ar' \in \allargs$,
  the claim $\forall i: [(\ar) \ileadstog \phi \lor (\ar') \ileadstog \phi]$ is not rejectable.
  In fact,
  % $\nexists \theta \in \Theta \suchthat \theta \notin C^\wsub$.
  $\Theta \wsub C^\wsub$.
  % Consider any ${\allleadstog} \notin C$, thus satisfying, for some $i$, $(\ar) \nileadstog \phi \land (\ar') \nileadstog \phi$.
  Consider any $θ \in \Theta$ and let us prove that $\theta \in \cup_{{\allleadstog} \in C} \powerset{\allleadstog}$.
  By definition of $\Theta$, $\theta^{-1}(P)$ may not contain both $((\ar), i)$ and $((\ar'), i)$, thus letting $t$ denote one of these pairs not in $\theta^{-1}(P)$ and defining $r = (t, \phi)$, pick any ${\allleadstog'}$ such that $θ \subset {\allleadstog'} \land t \in {\allleadstog'}$. We have $\theta \in \powerset{\allleadstog'}$ and ${\allleadstog'} \in C$.
\end{example}
\Cref{ex:disjR} shows a claim that is not rejectable, irrespective of the exposure protocol.
In particular, the example applies when $C$ does not hold (that is, when ${\allleadsto} \notin C$). We conclude that a claim may fail to hold but still be not rejectable.
It follows that non-rejection cannot be a sufficient condition for validating a claim.

@@

  Given any $θ \in \Theta$, define $l = \max_{\hist \suchthat \exists i \in I, p \in P \suchthat (i, \hist, p) \in θ} \card{\hist} + 1$ if $\exists p \in P \suchthat (i, \hist, p) \in θ$, and $l = 1$ otherwise. Pick any $\allleadstog$ such that $θ \subset {\allleadstog} \land (\ar)^l \ileadstog \phi$. We have
  % ${\allleadstog} \in S_θ \cap C ≠ \emptyset$.
  $\theta \in \powerset{\allleadstog}$ and ${\allleadstog} \in C$, thus, $\theta \in C^\wsub$.
\end{example}
\begin{example}[Disjunct with incompatible paths, not falsifiable]
  \label{ex:disj}
  Given $\ar \neq \ar' \in \allargs$, the claim appearing in \cref{ex:disjR}, namely,
  $\forall i: [(\ar) \ileadstog \phi \lor (\ar') \ileadstog \phi]$, is not falsifiable.
  Consider any ${\allleadstog} \notin C$, thus satisfying, for some $i$, $(\ar) \nileadstog \phi \land (\ar') \nileadstog \phi$.
  Consider any $θ \subseteq {\allleadstog}$ and let us prove that $\theta \in C^\wsub$.
  By definition of a general observation, $θ$ may not contain both $((\ar), i, \phi)$ and $((\ar'), i, \phi)$, thus letting $t$ denote one of these triples not in $θ$, pick any ${\allleadstog'}$ such that $θ \subset {\allleadstog'} \land t \in {\allleadstog'}$. We have 
  $\theta \in \powerset{\allleadstog'}$ and ${\allleadstog'} \in C$, thus, $\theta \in C^\wsub$.
  % ${\allleadstog'} \in S_θ \cap C ≠ \emptyset$.
\end{example}
\begin{example}[Partly falsifiable is still unfalsifiable]
  \label{ex:partlyU}
  Given $j \in I$, $\hist \in \allhists$ and $\ar \neq \ar' \in \allargs$,
  $\hist \ileadstog[j] \phi \land \forall i: [(\ar) \ileadstog \phi \lor (\ar') \ileadstog \phi]$ is not falsifiable.
  Consider any $\allleadstog$ satisfying $(\ar) \nileadstog \phi \land (\ar') \nileadstog \phi$, thus with ${\allleadstog} \notin C$.
  \Cref{ex:disj} concludes.
\end{example}
\begin{example}
  \label{ex:allParticular}
  Consider any $\ar$ and $\phi$. The claim $\forall i: \ar \ileadstostg \phi$ is falsifiable.
  Given any ${\allleadstog} \notin C$, consider any $i \suchthat \ar \nileadstostg \phi$ and any $\hist \suchthat \hist \nileadstog \phi \land \ar \in \histend$, and define $θ = \set{(i, \hist, {\allleadstog}(i, \hist) ≠ \phi)}$.
\end{example}
\begin{example}
  \label{ex:existsParticular}
  Given an infinite $I$, the claim $\exists i \suchthat \ar \ileadstostg \phi$ is not falsifiable.
  Given any ${\allleadstog} \notin C$, thus such that $\forall i: \ar \nileadstostg \phi$, any $θ \subseteq {\allleadstog}$ only has finitely many $(i, \hist, p ≠ \phi)$ thus $\theta \in C^\wsub$.
\end{example}

\begin{remark}
  The claim “For each glass, it will not break on a monday or not break on a tuesday, whatever happens to it.” is not falsifiable. But one can make it stronger and falsify it, by giving a general theory of how glasses break, proving that it holds, and that it is not compatible with the claim. I should think about this.
\end{remark}
\begin{remark}
  To be falsifiable, I need each time I exclude some relation $R$ to also exclude a subset $Cbar$ such that some $O \subseteq R$ satisfies $S_O$ is entirely in $Cbar$.
\end{remark}
\begin{theorem}
  A claim $C$ is falsifiable iff its complement $\Cbar = \allleadstos \setminus C$ satisfies:
  $\forall {\allleadstog} \in \Cbar, \exists \theta \subseteq {\allleadstog} \suchthat S_θ \subseteq \Cbar$, or equivalently:
  $\forall {\allleadstog} \in \Cbar, \exists O \in (\allhists)^I \suchthat S_{\restr{\allleadstog}{O_O}} \subseteq \Cbar$.
\end{theorem}
\begin{proof}
  The first phrasing follows from the definition of falsifiability.
  Also, the second phrasing implies the first one: pick $\theta = \restr{\allleadstog}{O_O}$.

>>>>

Given a pair $(I, \allargs)$, an observation $θ \in \Theta$ is a right unique finite subset of $(I × \allhists) × P$ such that $\forall i \in I, (i, \hist, p), (i, \histp, q) \in θ: \card{\hist} ≤ \card{\hist'} \implies \hist \subseteq \hist'$.
The set $\Theta$ contains the a priori possible observations, that is, those that can be considered possible when nothing is known about $\allleadsto$.
We say that an observation $\theta$ is compatible with a claim $C$ iff
$\theta \in C^\subseteq$, equivalently, iff
$\theta \in \cup_{{\allleadstog} \in C} \powerset{\allleadstog}$.
Thus, given ${\allleadstog} \in \allleadstos$, the set $\Theta \cap \powerset{\allleadstog}$ is the set of observations compatible with ${\allleadstog}$.

@@

Let the \emph{simple a priori observations} be defined as $\cup_{J \subseteq_\finite I} (\allhists)^J$, equivalently, as the right-unique finite relations over $I × \allhists$.
Given a simple a priori observation $O: J \mapsto \allhists$, we can define the a priori observation $O_O = \set{(i, \hist) \in J × \allhists \suchthat \forall i \in J: \hist \subseteq O(i)}$ as the completion of $O$ to all sub-paths in $O$.

@@

At our disposal is the procedure of empirical rejection, that consists in directly invalidating a claim by observing something that contradicts it.
\begin{definition}
  Given an exposure protocol $\allleadsto$, we say that $C$ is \emph{rejectable} iff
  $\Theta \cap \powerset{\allleadsto} \nsubseteq C^\wsub$.
\end{definition}

The following theorem verifies that when $C$ holds, it is not rejectable.
\begin{theorem}
  \label{th:ex-post}
  $\forall C \ni {\allleadsto}:
    \Theta \cap \powerset{\allleadsto} \subseteq C^\wsub$.
\end{theorem}
\begin{proof}
  This holds by definition of $C^\wsub = \cup_{{\allleadstog} \in C} \powerset{\allleadstog}$.
\end{proof}

The converse, however, does not hold.
\begin{example}[Disjunct with incompatible paths, not rejectable]
  \label{ex:disjR}
  Given $\ar \neq \ar' \in \allargs$,
  the claim $\forall i: [(\ar) \ileadstog \phi \lor (\ar') \ileadstog \phi]$ is not rejectable.
  In fact,
  $\Theta \wsub C^\wsub$.
  Consider any $θ \in \Theta$ and let us prove that $\theta \in \cup_{{\allleadstog} \in C} \powerset{\allleadstog}$.
  By definition of $\Theta$, $\theta^{-1}(P)$ may not contain both $((\ar), i)$ and $((\ar'), i)$, thus letting $t$ denote one of these pairs not in $\theta^{-1}(P)$ and defining $r = (t, \phi)$, pick any ${\allleadstog'}$ such that $θ \subset {\allleadstog'} \land t \in {\allleadstog'}$. We have $\theta \in \powerset{\allleadstog'}$ and ${\allleadstog'} \in C$.
\end{example}
\Cref{ex:disjR} shows a claim that is not rejectable, irrespective of the exposure protocol.
In particular, the example applies when $C$ does not hold (that is, when ${\allleadsto} \notin C$). We conclude that a claim may fail to hold but still be not rejectable.
It follows that non-rejection cannot be a sufficient condition for validating a claim.

@@

  Given any $θ \in \Theta$, define $l = \max_{\hist \suchthat \exists i \in I, p \in P \suchthat (i, \hist, p) \in θ} \card{\hist} + 1$ if $\exists p \in P \suchthat (i, \hist, p) \in θ$, and $l = 1$ otherwise. Pick any $\allleadstog$ such that $θ \subset {\allleadstog} \land (\ar)^l \ileadstog \phi$. We have
  $\theta \in \powerset{\allleadstog}$ and ${\allleadstog} \in C$, thus, $\theta \in C^\wsub$.
\end{example}
\begin{example}[Disjunct with incompatible paths, not falsifiable]
  \label{ex:disj}
  Given $\ar \neq \ar' \in \allargs$, the claim appearing in \cref{ex:disjR}, namely,
  $\forall i: [(\ar) \ileadstog \phi \lor (\ar') \ileadstog \phi]$, is not falsifiable.
  Consider any ${\allleadstog} \notin C$, thus satisfying, for some $i$, $(\ar) \nileadstog \phi \land (\ar') \nileadstog \phi$.
  Consider any $θ \subseteq {\allleadstog}$ and let us prove that $\theta \in C^\wsub$.
  By definition of a general observation, $θ$ may not contain both $((\ar), i, \phi)$ and $((\ar'), i, \phi)$, thus letting $t$ denote one of these triples not in $θ$, pick any ${\allleadstog'}$ such that $θ \subset {\allleadstog'} \land t \in {\allleadstog'}$. We have 
  $\theta \in \powerset{\allleadstog'}$ and ${\allleadstog'} \in C$, thus, $\theta \in C^\wsub$.
\end{example}
\begin{example}[Partly falsifiable is still unfalsifiable]
  \label{ex:partlyU}
  Given $j \in I$, $\hist \in \allhists$ and $\ar \neq \ar' \in \allargs$,
  $\hist \ileadstog[j] \phi \land \forall i: [(\ar) \ileadstog \phi \lor (\ar') \ileadstog \phi]$ is not falsifiable.
  Consider any $\allleadstog$ satisfying $(\ar) \nileadstog \phi \land (\ar') \nileadstog \phi$, thus with ${\allleadstog} \notin C$.
  \Cref{ex:disj} concludes.
\end{example}
\begin{example}
  \label{ex:allParticular}
  Consider any $\ar$ and $\phi$. The claim $\forall i: \ar \ileadstostg \phi$ is falsifiable.
  Given any ${\allleadstog} \notin C$, consider any $i \suchthat \ar \nileadstostg \phi$ and any $\hist \suchthat \hist \nileadstog \phi \land \ar \in \histend$, and define $θ = \set{(i, \hist, {\allleadstog}(i, \hist) ≠ \phi)}$.
\end{example}
\begin{example}
  \label{ex:existsParticular}
  Given an infinite $I$, the claim $\exists i \suchthat \ar \ileadstostg \phi$ is not falsifiable.
  Given any ${\allleadstog} \notin C$, thus such that $\forall i: \ar \nileadstostg \phi$, any $θ \subseteq {\allleadstog}$ only has finitely many $(i, \hist, p ≠ \phi)$ thus $\theta \in C^\wsub$.
\end{example}

\begin{remark}
  The claim “For each glass, it will not break on a monday or not break on a tuesday, whatever happens to it.” is not falsifiable. But one can make it stronger and falsify it, by giving a general theory of how glasses break, proving that it holds, and that it is not compatible with the claim. I should think about this.
\end{remark}
Let $S_θ = \set{{\allleadstog} \in \allleadstos \suchthat θ \subseteq {\allleadstog}}$ denote the possibilities for $\allleadsto$ compatible with the observations.
Given $O$ and ${\allleadstog}$, it follows that $S_{\restr{\allleadstog}{O_O}}$ denotes the “leads to” relations that complete $O_O$ using $\allleadstog$.
\begin{remark}
  To be falsifiable, I need each time I exclude some relation $R$ to also exclude a subset $Cbar$ such that some $O \subseteq R$ satisfies $S_O$ is entirely in $Cbar$.
\end{remark}
\begin{theorem}
  A claim $C$ is falsifiable iff its complement $\Cbar = \allleadstos \setminus C$ satisfies:
  $\forall {\allleadstog} \in \Cbar, \exists \theta \subseteq {\allleadstog} \suchthat S_θ \subseteq \Cbar$, or equivalently:
  $\forall {\allleadstog} \in \Cbar, \exists O \in (\allhists)^I \suchthat S_{\restr{\allleadstog}{O_O}} \subseteq \Cbar$.
\end{theorem}
\begin{proof}
  The first phrasing follows from the definition of falsifiability.
  Also, the second phrasing implies the first one: pick $\theta = \restr{\allleadstog}{O_O}$.
